---
title: "Homework 6"
subtitle: "4375 Machine Learning with Dr. Mazidi"
author: "Parker Whitehead"
date: "10/2/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Problem 1: Comparison with Linear Regression

### Step 1. Load Auto data and make train/test split

Using the Auto data in package ISLR, set seed to 1234 and divide into 75% train, 25% test

```{r}
# your code here
library(ISLR)
data(Auto)
set.seed(1234)

i <- sample(1:nrow(Auto),nrow(Auto)*.75,replace=FALSE)
train <- Auto[i,]
test <- Auto[-i,]
```

### Step 2. Build  linear regression model

Build a linear regression model on the train data, with mpg as the target, and cylinders, displacement, and horsepower as the predictors.  Output a summary of the model and plot the model to look at the residuals plots.

```{r}
lm1 <- lm(mpg~cylinders+displacement+horsepower, data=train)
summary(lm1) 
par(mfrow=c(2,2))
plot(lm1)
par(mfrow=c(1,1))
```

### Step 3. Evaluate on the test data

Evaluate the model on the test data. Output correlation and mse.

```{r}
# your code here
pred <- predict(lm1, newdata=test)
correlation <- cor(pred,test$mpg)
mse <- mean((pred-test$mpg)^2)
print(paste("correlation: ",correlation))
print(paste("mse: ",mse))
```

### Step 4. Try knn

Use knnreg() in library caret to fit the training data. Use the default k=1. Output your correlation and mse.

```{r}
# your code here
library(caret)
fit <- knnreg(train[,2:4],train[,1],k=1)
pred <- predict(fit, test[,2:4])
correlation <- cor(pred,test$mpg)
mse <- mean((pred-test$mpg)^2)
print(paste("correlation: ",correlation))
print(paste("mse: ",mse))
```

### Step 5. Analysis

a.	Compare correlation metric that each algorithm achieved. Your commentary here:
  Although both correlation metrics were within an acceptable range, the correlation for the knn model was better.

b.	Compare the mse metric that each algorithm achieved. Your commentary here:
  Similar to what we found in correlation, the mse is acceptable for both, but better for the knn model.

c.	Why do you think that the mse metric was so different compared to the correlation metric?  Your commentary here:
  In my case, the change in mse and correlation were not that different. Both equally improved by a moderate margin.

d.	Why do you think that kNN outperformed linear regresssion on this data? In your 2-3 sentence explanation, discuss bias of the algorithms. Your commentary here:
  This could be a result of the model being clumped together in groups rather than fit to an orderly line. Linear regression struggles with models that have data that is spread across the Y axis. This is not an issue with knn, as it does not create a line, and instead compares items to what they are near. This avoids the linear bias.


# Problem 2: Comparison with Logistic Regression

### Step 1.  Load Breast Cancer data, create regular and small factors, and divide into train/test

Using the BreastCancer data in package mlbench, create factor columns Cell.small and Cell.regular as we did in the last homework. Set seed to 1234 and divide into 75% train, 25% test. 

*Advice*: use different names for test/train so that when you run parts of  your script over and over the names don’t collide.

```{r}
library(mlbench)
data(BreastCancer)
BreastCancer$Cell.small <- FALSE
BreastCancer$Cell.small <- ifelse(BreastCancer$Cell.size==1,1,0)
BreastCancer$Cell.small <- factor(BreastCancer$Cell.small)

BreastCancer$Cell.regular <- FALSE
BreastCancer$Cell.regular <- ifelse(BreastCancer$Cell.shape==1,1,0)
BreastCancer$Cell.regular <- factor(BreastCancer$Cell.regular)

set.seed(1234)
i <- sample(1:nrow(BreastCancer),nrow(BreastCancer)*.75, replace = FALSE)
train2 <- BreastCancer[i,]
test2 <- BreastCancer[-i,]
```


### Step 2. Build logistic regression model

Build a logistic regression model with Class as the target and Cell.small and Cell.regular as the predictors. Output a summary of the model. 

```{r}
glm1 <-glm(Class~Cell.small+Cell.regular, data=train2, family=binomial)
summary(glm1)
```

### Step 3. Evaluate on the test data

Evaluate the model on the test data. Output accuracy and a table (or confusion matrix).

```{r}
probs2 <- predict(glm1,newdata=test2,type="response")
pred2 <- ifelse(probs2>.5,"malignant","benign")
acc <- mean(pred2==test2$Class)
print(paste("accuracy: ",acc))
table(pred2,test2$Class)

```
 
### Step 4. Try knn

Use the knn() function in package class to use the same target and predictors as step 2. Output accuracy and a table of results for knn. 

```{r}
# your code here
library(class)
knn2 <- knn(train=train2[,12:13],test=test2[,12:13],cl=train2$Class)
acc2 <- mean(knn2==test2$Class)
print(paste("accuracy: ",acc2))
table(knn2,test2$Class)
```

### Step 5. Try knn on original predictors

Run kNN using predictor columns 2-6, 8-10, using default k=1.  Output accuracy and a table of results.

Compare the results from step 5 above to a model which uses all the predictors. Provide some analysis on why you see these results:

```{r}
# your code here
library(class)
train_pred <- train2[,c(2:6,8:10)]
test_pred <- test2[,c(2:6,8:10)]
train_tar <- train2[,11]
test_tar <- test2[,11]
knn2 <- knn(train=train_pred,test=test_pred,cl=train_tar,k=1)
acc2 <- mean(knn2==test_tar)
print(paste("accuracy: ",acc2))
table(knn2,test_tar)
```

### Step 6. Try logistic regression on original predictors

Run logistic regression using predictor columns 2-6, 8-10.  Output accuracy and a table of results.

Compare the results from the logistic regression and knn algorithms using all predictors except column 7 in the steps above. Provide some analysis on why you see these results:

Incredibly similar results are seen between the two sets. This is most likely due to the fact that so many predictors are used, and that there are only 2 categories to be made. 


```{r}
glm2 <-glm(Class~Cl.thickness+Cell.size+Cell.shape+Marg.adhesion+Epith.c.size+Bl.cromatin+Normal.nucleoli+Mitoses, data=train2, family=binomial)
probs3 <- predict(glm2,newdata=test2,type="response")
pred3 <- ifelse(probs3>.5,"malignant","benign")
acc <- mean(pred3==test2$Class)
print(paste("accuracy: ",acc))
table(pred3,test2$Class)
```








