---
title: "Homework 1"
subtitle: "4375 Machine Learning with Dr. Mazidi"
author: "Parker Whitehead"
date: "8/30/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This homework has two parts:

* Part 1 uses R for data exploration
* Part 2 uses C++ for data exploration

---

This homework is worth 100 points, 50 points each for Part 1 and Part 2. 

---

# Part 1: RStudio Data Exploration

**Instructions:** Follow the instructions for the 10 parts below. If the step asks you to make an observation or comment, write your answer in the white space above the gray code box for that step. 

## Step 1: Load and explore the data

* load library MASS (install at console, not in code)
* load the Boston dataframe using data(Boston)
* use str() on the data
* type ?Boston at the console
* Write 2-3 sentences about the data set below

Your commentary here:
The Boston data set contains numerous data points on the suburbs of Boston. 
It contains 506 rows and 14 variables, or columns.
The 14 variables cover many statistics, from per capita crime rate to full-value property tax rate per $10k.

```{r}
# step 1 code
library(MASS)
data(Boston)
str(Boston)
?Boston
```

## Step 2: More data exploration

Use R commands to:

* display the first few rows
* display the last two rows
* display row 5
* display the first few rows of column 1 by combining head() and using indexing
* display the column names

```{r}
# step 2 code
head(Boston) #first few
tail(Boston,n=2) #last 2
Boston[5,] #row 5
head(Boston[,1]) #first few rows of col 1
colnames(Boston) #col names
```

## Step 3: More data exploration

For the crime column, show:

* the mean
* the median
* the range

```{r}
# step 3 code
mean(Boston$crim)
median(Boston$crim)
range(Boston$crim)
```

## Step 4: Data visualization

Create a histogram of the crime column, with an appropriate main heading. In the space below, state your conclusions about the crime variable:

Your commentary here:
The vast majority of all crime rates appear between 0-10 per capita falling off hard past 20 per capita.

```{r}
# step 4 code
perCapitaCrimeRate <- Boston$crim #placed into var so that the main heading is appropriate.
hist(perCapitaCrimeRate)

```

## Step 5: Finding correlations

Use the cor() function to see if there is a correlation between crime and median home value. In the space below, write a sentence or two on what this value might mean. Also write about whether or not the crime column might be useful to predict median home value.

Your commentary here:
There does not seem to be a strong coorelation between medv and crim, as the abs value returned from cor() is significantly less than .7
Using plot(), we can see that medv greatly varies in areas with low crime rates. Considering this and the correlation value from cor(), we can determine that crime rate might not be sufficient in predicting median home value, especially if the crime rate is low.

```{r}
# step 5 code
b_cvm <- Boston[,c(1,14)] #Storing medv and crim into b_cvm, which stands for "boston crim vs medv"
cor(b_cvm)
```

## Step 6: Finding potential correlations

Create a plot showing the median value on the y axis and number of rooms on the x axis. Create appropriate main, x and y labels, change the point color and style. [Reference for plots(http://www.statmethods.net/advgraphs/parameters.html) 

Use the cor() function to quantify the correlation between these two variables. Write a sentence or two summarizing what the graph and correlation tell you about these 2 variables.

Your commentary here:
Outside of a few anomalies, there seems to be a much greater correlation between rooms and median value than there was between crime and median value. Specifically, there is a fair linear correlation between the two. 
```{r}
# step 6 code
b_mvr <- Boston[,c(6,14)] #Storing medv and crim into b_cvm, which stands for "boston medv vs number of rooms"
plot(b_mvr,main="Median House Values compared to Number of Rooms", xlab="# of Rooms", ylab="Median House Value",pch=19,col="red")
cor(b_mvr)
```

## Step 7: Evaluating potential predictors

Use R functions to determine if variable chas is a factor. Plot median value on the y axis and chas on the x axis. Make chas a factor and plot again.

Comment on the difference in meaning of the two graphs. Look back the description of the Boston data set you got with the ?Boston command to interpret the meaning of 0 and 1.  

Your commentary here:
By factoring chas, we can allow the plot() function to provide move clear information on chas as a whole. Chas actually represents whether or not a tract bounds river (1 if so, 0 if not). From the factored plot, it is clear to see that houses whose tract bounds the river tend to have a higher medv.

```{r}
# step 7 code
b_mvch <- Boston[,c(4,14)] #storing chas and medv into b_mvch, which stands for "boston medv vs chas"
plot(b_mvch)
is.factor(b_mvch$chas)
b_mvch$chas <- factor(b_mvch$chas) #since chas isn't a factor, b_mvch$chas is overwritten as a factor.
plot(b_mvch)
```

## Step 8: Evaluating potential predictors

Explore the rad variable. What kind of variable is rad? What information do you get about this variable with the summary() function? Does the unique() function give you additional information? Use the sum() function to determine how many neighborhoods have rad equal to 24. Use R code to determine what percentage this is of the neighborhoods.

Your commentary here:
The variable rad is an integer. Summary gives us lots of different observations of rad, such as Mean, Median, Max, etc. The unique function displays all unique values within rad. By using the sum function, I was able to determine that there are 132 houses with rad==24. This makes up ~26% of the neighborhoods.

```{r}
# step 8 code
typeof(Boston$rad)
summary(Boston$rad)
unique(Boston$rad)
radSum <- Boston$rad==24
radSum <- sum(radSum)
percentage <- radSum/length(Boston$rad)
```

## Step 9: Adding a new potential predictor

Create a new variable called “far” using the ifelse() function that is TRUE if rad is 24 and FALSE otherwise. Make the variable a factor. Plot far and medv. What does the graph tell you? 

Your commentary here:
The plotting displays that there is that for rad values of 24, the medv tends to be lower than for other values of rad.

```{r}
# step 9 code
Boston$far <- ifelse(Boston$rad==24,TRUE,FALSE)
Boston$far <- factor(Boston$far) #Far is a factor that is true if the rad value is == 24.
b_fvm <- Boston[,c(15,14)] #b_fvm means boston_far vs medv.
plot(b_fvm)
```

## Step 10: Data exploration

* Create a summary of Boston just for columns 1, 6, 13 and 14 (crim, rm, lstat, medv)
* Use the which.max() function to find the neighborhood with the highest median value. See p. 176 in the pdf
* Display that row from the data set, but only columns 1, 6, 13 and 14
* Write a few sentences comparing this neighborhood and the city as a whole in terms of: crime, number of rooms, lower economic percent, median value.

Your commentary here:
The highest median value neighborhood hood also has the minimum lower status value out of the entire city, which makes sense given the medv of the neighborhood. This neighborhood also experiences less crime than the overall mean, but is still quite a bit higher than the city's median. Suprisingly, however, it's number of rooms hovers fairly close to the city's mean.

```{r}
# step 10 code
sum_sample <-Boston[,c(1,6,13,14)]
summary(sum_sample)
highest_median_value_index <- which.max(Boston$medv)
str(sum_sample[highest_median_value_index,])
```

# Part 2: C++

In this course we will get some experience writing machine learning algorithms from scratch in C++, and comparing performance to R. Part 2 of Homework 1 is designed to lay the foundation for writing custom machine learning algorithms in C++. 
	
To complete Part 2, first you will read in the Boston.csv file which just contains columns rm and medv. 
	
--- 

In the C++ IDE of your choice:
	
1	Read the csv file (now reduced to 2 columns) into 2 vectors of the appropriate type.

2	Write the following functions: 

*	a function to find the sum of a numeric vector

*	a function to find the mean of a numeric vector

*	a function to find the median of a numeric vector

*	a function to find the range of a numeric vector

*	a function to compute covariance between rm and medv (see formula on p. 74 of pdf)

*	a function to compute correlation between rm and medv (see formula on p. 74 of pdf); Hint: sigma of a vector can be calculated as the square root of variance(v, v)

3	Call the functions described in a-d for rm and for medv. Call the covariance and correlation functions. Print results for each function.







